#  2: Regression and model validation


```{r}
date()
```

First reading in the data that got wrangled before and exploring the structure and the dimensions of the data frame. 
```{r}
library(tidyverse)
learning2014 <- read.csv("data/learning2014.csv")

dim(learning2014)
# gives the amount of observations and variables in the data frame 
str(learning2014)
# gives the structure of the data frames, including the variable classes of each variable
```

This data set consists of 166 observations (rows) of 7 variables (columns), collected from a 2014 international survey of Approaches to Learning. The variables in the data frame are the gender and age of the student, the points the student got in the course exam, and mean values of questions related to a) students attitude towards statistics, b) to deep, c) surface and d) strategic learning, each measured on the Likert scale (1-5). 


To view the data graphically, a scatter matrix of the variables is plotted and shown below.

```{r}
# access the GGally and ggplot2 libraries
library(GGally)
library(ggplot2)

# create a plot matrix with ggpairs()
p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p
```


